Another Look at Regularized Discriminant Analysis for High-Dimensional Classification

### 2013-04-21 reports/paper-rda.Rnw
Addressed TODO items in footnotes.

Here, we propose an improvement to the popular regularized discriminant
analysis classifier of \cite{Friedman:1989tm}, who has proposed pooling of the
covariance-matrix estimators from linear and quadratic discriminant analysis and
has additionally incorporated a covariance-matrix shrinkage term to ensure that
the classifier is well-posed. \citeauthor{Friedman:1989tm}'s classifier can
attain improved classification accuracy, compared to linear and quadratic
discriminant analysis when $p < N$. However, in its original form,
\citeauthor{Friedman:1989tm}'s classifier is impractical when $p \gg N$ because
the decision rule requires the calculation of the inverses and determinants of
large $p \times p$ covariance-matrix estimators for each class. Furthermore,
\citeauthor{Friedman:1989tm}'s classifier has two tuning parameters that are
selected via cross-validation, which adds a substantial computational burden to
the model-selection process. \citeauthor{Friedman:1989tm}'s classifier is
ill-advised on modern data sets because the model selection requires $O(Kp^3)$
calculations for each tuning-parameter pair.

@@

\subsection{Improvements to \citeauthor{Friedman:1989tm}'s Classifier}

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed to increase eigenvalues (approximately) equal to 0. Here, we define the
eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\Sigma}_k = \alpha_k \widehat{\Sigma}_k(\lambda) + \gamma I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. The choice of $\alpha_k$ allows for a flexible formulation of
covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1, \ldots,
K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}. Similarly,
if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form similar to
\eqref{eq:sig-rda}. Now, substituting \eqref{eq:hdrda-cov} into \eqref{eq:qda},
we define the high-dimensional regularized discriminant analysis classifier
\begin{align}
	D_{HDRDA}(x) = \argmin_{k}  (x - \xbar_k)^{T}\tilde{\Sigma}_k^{+}(x - \xbar_k)  + \log |\tilde{\Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\Sigma}_k$ is nonsingular such that
$\tilde{\Sigma}_k^{-1}$ can be substituted for $\tilde{\Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\Sigma}_k$.

@@

Next, we demonstrate that our proposed classifier in \eqref{eq:hdrda} lends
itself to a more efficient calculation and a rapid model selection of the
parameters $\lambda$ and $\gamma$\footnote{TODO: Briefly mention the updating
  trick employed in the R package}. We decompose \eqref{eq:hdrda} into a sum of
two components, where the first summand consists of matrix operations applied to
low-dimensional matrices and the second summand corresponds to the null space of
$\widehat{\Sigma}$ in \eqref{eq:pooled-cov}. We show that the matrix
operations performed on the null space of $\widehat{\Sigma}$ yield constant
quadratic forms across all classes and can be omitted. For $p \gg N$, the
constant component involves determinants and inverses of high-dimensional
matrices, and by ignoring these calculations, a substantial reduction in
computational costs is achieved relative to Friedman's classifier in
\eqref{eq:rda}\footnote{TODO: Add a brief comparison of the computations of
  \emph{RDA} and \emph{HDRDA} in terms of Big Oh notation}. Furthermore, a
byproduct is that adjustments to the associated eigenvalues have no effect on
\eqref{eq:hdrda}. Lastly, we utilize the singular value decomposition to
calculate efficiently the eigenvalue decomposition of $\widehat{\Sigma}$,
further reducing the computational costs of our proposed classifier.

@@

\subsection{Model Selection}

Thus far, we have shown that our classifier in \eqref{eq:hdrda} is invariant to
the term $U_2$, thus yielding an equivalent classifier in
\eqref{eq:hdrda-decomposed} with a substantial reduction in the computational
complexity. Here, we provide two\footnote{TODO: Two? Perhaps, I mean one and
  then the Fast SVD below. If so, clarify this.} additional results that further
improve the computational performance of our classifier. First, we demonstrate
that the inverse and determinants of $W_k$ can be written so that they are
performed on $n_k \times n_k$ matrices rather than the larger $q \times q$
matrices, resulting in further improvements to the computational efficiency.

\begin{proposition}\label{proposition:rda-W_k}
Let $W_k$ be defined as above $(k = 1, \ldots, K)$. Then, $|W_k| = |
\Gamma_k| |Q_k|$ and
	\begin{align}
		W_k^{-1} &= \Gamma_k^{-1} - \alpha_k(1 - \lambda) \Gamma_k^{-1} U_1^{T} X_k^{T} Q_k^{-1} X_k U_1 \Gamma_k^{-1},\label{eq:W_k_inv}
	\end{align}
where $Q_k = I_{n_k} + \alpha_k(1- \lambda) X_k U_1 
\Gamma_k^{-1} U_1^{T} X_k^{T}$ and $\Gamma_k = \alpha_k \lambda D_q +
\gamma I_q$.
\end{proposition}
\begin{proof}
To calculate $|W_k|$, we apply Theorem 18.1.1 from \cite{Harville:2008wja},
which states that $|A + B T C| = |A| |T| |T^{-1} +
C A^{-1} B|$. Thus, setting $A = \Gamma_k$, $B =
\alpha_k (1 - \lambda) U_1^{T} X_k^{T}$, $T = I_{n_k}$, and $C =
X_k U_1$, we have $|W_k| = |\Gamma_k| |Q_k|$. Also,
\eqref{eq:W_k_inv} follows from the well-known Sherman-Woodbury formula
\citep[Theorem 18.2.8]{Harville:2008wja}.
\end{proof}

@@

In this section, we describe four high-dimensional microarray data sets and
compare our proposed classifier with five classifiers recently proposed for
small-sample, high-dimensional data. Our competitors include diagonal linear
discriminant analysis from \cite{Dudoit:2002ev}, shrunken centroids regularized
discriminant analysis from \cite{Guo:2007te}, and penalized linear discriminant
analysis from \cite{Witten:2011kc}. We also consider two modifications of
diagonal linear discriminant analysis from \cite{Tong:2012hw} and
\cite{Pang:2009ik}, where the former employs an improved mean estimator and the
latter utilizes an improved variance estimator. Below, we refer to each
classifier by the first author's surname.

We evaluated the classification performance of each classifier by randomly
partitioning the data such that $2/3$ of the observations were allocated as
training data and the remaining $1/3$ of the observations were allocated as a
test data set. For each random partition, we applied the feature-selection
method from \cite{Dudoit:2002ev} to the training data to choose the $d$ features
having the largest ratio of their between-group to within-group sums of squares.
We then filtered the test data set with the corresponding $d$ features. After
training each classifier from the filtered training data, we classified the
filtered test data sets and computed the proportion of mislabeled test
observations to estimate the classification error rate for each
classifier. Repeating this process 1000 times, we computed the average of the
error-rate estimates for each classifier. We remark that our simulation design
avoids the selection bias discussed in \cite{Ambroise:2002fa}. We did not
include \citeauthor{Friedman:1989tm}'s classifier in our studies to avoid its
excessive computations. In Table \ref{tab:data-summary}, we summarize the
attributes of each microarray data set.

We used version 3.0.0 of the open-source statistical software {\tt R} for the
classification study. For each classifier's {\tt R} implementation, we used the
default settings with the exception that we explicitly set prior probabilities
as equal. For our classifier, we set $\alpha_k = 1$ $(k = 1, \ldots, K)$, so
that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed had a form
resembling \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21
equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
10-fold cross-validation \citep{Hastie:2008dt}.

@@

In our simulation study, we compared our classifier to several recently proposed
classifiers, including the Dudoit, Pang, Tong, and Guo classifiers, each of
which assumes independence of the features conditional on class
membership. Despite the rapid computational performance of diagonal classifiers
and the reduction in the number of parameters to estimate, \cite{Mai:2012bf} and
\cite{Fan:2012iq} have noted that diagonal classifiers can often yield inferior
classification performance to other classification methods. We have shown that
our proposed classifier often yields superior classification accuracy to the
diagonal classifiers considered, confirming the assertions of \cite{Mai:2012bf}
and \cite{Fan:2012iq}. Furthermore, the implementation of our classifier in the
{\tt sparsediscrim} R package available on CRAN was comparable to that of the
diagonal classifiers in terms of computational runtime. Moreover, when
preliminary variable selection was applied to the data in our classification
study, the time to perform model selection with our proposed classifier was
nearly instantaneous.

>>>>

Here, we propose an improvement to the popular regularized discriminant
analysis classifier of \cite{Friedman:1989tm}, who has proposed pooling of the
covariance-matrix estimators from linear and quadratic discriminant analysis and
has additionally incorporated a covariance-matrix shrinkage term to ensure that
the classifier is well-posed. \citeauthor{Friedman:1989tm}'s classifier can
attain improved classification accuracy, compared to linear and quadratic
discriminant analysis when $p < N$. However, in its original form,
\citeauthor{Friedman:1989tm}'s classifier is impractical when $p \gg N$ because
the decision rule requires the calculation of the inverses and determinants of
large $p \times p$ covariance-matrix estimators for each class. Furthermore,
\citeauthor{Friedman:1989tm}'s classifier has two tuning parameters that are
selected via cross-validation, which adds a substantial computational burden to
the model-selection process.

@@

\subsection{Model Specification}

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed to increase eigenvalues (approximately) equal to 0. Here, we define the
eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\Sigma}_k = \alpha_k \widehat{\Sigma}_k(\lambda) + \gamma I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. The choice of $\alpha_k$ allows for a flexible formulation of
covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1, \ldots,
K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}. Similarly,
if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form similar to
\eqref{eq:sig-rda}. Now, substituting \eqref{eq:hdrda-cov} into \eqref{eq:qda},
we define the high-dimensional regularized discriminant analysis classifier
\begin{align}
	D_{HDRDA}(x) = \argmin_{k}  (x - \xbar_k)^{T}\tilde{\Sigma}_k^{+}(x - \xbar_k)  + \log |\tilde{\Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\Sigma}_k$ is nonsingular such that
$\tilde{\Sigma}_k^{-1}$ can be substituted for $\tilde{\Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\Sigma}_k|$ equal
to the product of the positive eigenvalues of $\tilde{\Sigma}_k$. Following
\cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$ from a grid of
candidate models via cross-validation \citep{Hastie:2008dt}. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function in the {\tt
  sparsediscrim} R package, available on CRAN.

@@

Next, we demonstrate that our proposed classifier in \eqref{eq:hdrda} lends
itself to a more efficient calculation. We decompose \eqref{eq:hdrda} into a sum
of two components, where the first summand consists of matrix operations applied
to low-dimensional matrices and the second summand corresponds to the null space
of $\widehat{\Sigma}$ in \eqref{eq:pooled-cov}. We show that the matrix
operations performed on the null space of $\widehat{\Sigma}$ yield constant
quadratic forms across all classes and can be omitted. For $p \gg N$, the
constant component involves determinants and inverses of high-dimensional
matrices, and by ignoring these calculations, a substantial reduction in
computational costs is achieved relative to Friedman's classifier in
\eqref{eq:rda}. Furthermore, a byproduct is that adjustments to the associated
eigenvalues have no effect on \eqref{eq:hdrda}. Lastly, we utilize the singular
value decomposition to calculate efficiently the eigenvalue decomposition of
$\widehat{\Sigma}$, further reducing the computational costs of our proposed
classifier.

@@

Thus far, we have shown that our classifier in \eqref{eq:hdrda} is invariant to
the term $U_2$, thus yielding an equivalent classifier in
\eqref{eq:hdrda-decomposed} with a substantial reduction in the computational
complexity. Here, we demonstrate that the computational efficiency in
calculating the inverse and determinants of $W_k$ can be further improved via
standard matrix operations.

\begin{proposition}\label{proposition:hdrda-W_k}
Let $W_k$ be defined as above $(k = 1, \ldots, K)$. Then, $|W_k| = |
\Gamma_k| |Q_k|$ and
	\begin{align}
		W_k^{-1} &= \Gamma_k^{-1} - \alpha_k(1 - \lambda) \Gamma_k^{-1} U_1^{T} X_k^{T} Q_k^{-1} X_k U_1 \Gamma_k^{-1},\label{eq:W_k_inv}
	\end{align}
where $Q_k = I_{n_k} + \alpha_k(1 - \lambda) X_k U_1 
\Gamma_k^{-1} U_1^{T} X_k^{T}$ and $\Gamma_k = \alpha_k \lambda D_q +
\gamma I_q$.
\end{proposition}
\begin{proof}
To calculate $|W_k|$, we apply Theorem 18.1.1 from \cite{Harville:2008wja},
which states that $|A + B T C| = |A| |T| |T^{-1} +
C A^{-1} B|$. Thus, setting $A = \Gamma_k$, $B =
\alpha_k (1 - \lambda) U_1^{T} X_k^{T}$, $T = I_{n_k}$, and $C =
X_k U_1$, we have $|W_k| = |\Gamma_k| |Q_k|$. Also,
\eqref{eq:W_k_inv} follows from the well-known Sherman-Woodbury formula
\citep[Theorem 18.2.8]{Harville:2008wja}.
\end{proof}

Conveniently, we see that the inverse and determinant of $W_k$ are each in terms
of the inverse and determinant of $Q_k \in \mathbb{R}_{n_k \times n_k}$,
respectively, as well as the diagonal matrix $\Gamma_k \in \mathbb{R}_{q \times
  q}$. This formulation is especially advantageous when the number of classes
$K$ is large because $n_k$ is then potentially much smaller than $q$. The
expressions given in Proposition \ref{proposition:hdrda-W_k} also expedite the
selection of $\lambda$ and $\gamma$ via cross-validation. The most
time-consuming matrix operations involved in computing $W_k^{-1}$ and $|W_k|$ is
$X_k U_1 \in \mathbb{R}_{n_k \times q}$, which is independent of $\lambda$ and
$\gamma$. Hence, we need calculate $X_k U_1$ only once for each cross-validation
fold. The subsequent operations in calculating $W_k^{-1}$ and $|W_K|$ can be
simply updated for different pairs of $\lambda$ and $\gamma$.

@@

In this section, we describe four high-dimensional microarray data sets and
compare our proposed classifier with five classifiers recently proposed for
small-sample, high-dimensional data. In Table \ref{tab:data-summary}, we
summarize the four microarray data sets. Our competitors include diagonal linear
discriminant analysis from \cite{Dudoit:2002ev}, shrunken centroids regularized
discriminant analysis from \cite{Guo:2007te}, and penalized linear discriminant
analysis from \cite{Witten:2011kc}. We also consider two modifications of
diagonal linear discriminant analysis from \cite{Tong:2012hw} and
\cite{Pang:2009ik}, where the former employs an improved mean estimator and the
latter utilizes an improved variance estimator. Below, we refer to each
classifier by the first author's surname.

We evaluated the classification performance of each classifier by randomly
partitioning the data such that $2/3$ of the observations were allocated as
training data and the remaining $1/3$ of the observations were allocated as a
test data set. For each random partition, we applied the feature-selection
method from \cite{Dudoit:2002ev} to the training data to choose the $d$ features
having the largest ratio of their between-group to within-group sums of squares.
We then filtered the test data set with the corresponding $d$ features. After
training each classifier from the filtered training data, we classified the
filtered test data sets and computed the proportion of mislabeled test
observations to estimate the classification error rate for each
classifier. Repeating this process 1000 times, we computed the average of the
error-rate estimates for each classifier. We remark that our simulation design
avoids the selection bias discussed in \cite{Ambroise:2002fa}. We did not
include \citeauthor{Friedman:1989tm}'s classifier in our studies to avoid its
excessive computations.

We used version 3.0.0 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively, on
CRAN. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule in our {\tt R} code. We also utilized
our implementations of the Dudoit, Pang, and Tong classifiers available in the
{\tt sparsediscrim} package. We used the default settings for each classifier
with the exception that we explicitly set prior probabilities as equal, if the
option was available. For our classifier, we set $\alpha_k = 1$ $(k = 1, \ldots,
K)$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed had a
form resembling \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid
of 21 equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
10-fold cross-validation \citep{Hastie:2008dt}.

@@

In our simulation study, we compared our classifier to several recently proposed
classifiers, including the Dudoit, Pang, Tong, and Guo classifiers, each of
which assumes independence of the features conditional on class
membership. Despite the rapid computational performance of diagonal classifiers
and the reduction in the number of parameters to estimate, \cite{Mai:2012bf} and
\cite{Fan:2012iq} have noted that diagonal classifiers can often yield inferior
classification performance to other classification methods. We have shown that
our proposed classifier often yields superior classification accuracy to the
diagonal classifiers considered, confirming the assertions of \cite{Mai:2012bf}
and \cite{Fan:2012iq}. Furthermore, the implementation of our classifier in the
{\tt sparsediscrim} R package was comparable to that of the diagonal classifiers
in terms of computational runtime. Moreover, when preliminary variable selection
was applied to the data in our classification study, the time to perform model
selection with our proposed classifier was nearly instantaneous.
